Inferring the exoplanet radius distribution from a noisy, incomplete catalog

### 2013-11-26 document/ms.tex
adding general discussion

>>>>

\section{Hierarchical inference}

If we assume that the exoplanetary systems that we have observed are drawn (in
noisy and biased way) from some global distribution then it would be
interesting to infer the shape of this global distribution conditioned on all
the available data and our understanding of survey selection effects.
This problem is called the \emph{hierarchical inference problem}.
In theory, this problem is no different than a standard inference problem and
the same algorithms might be applicable but in general, this class of problems
refer to the cases where forward modelling directly to the raw data is
computationally intractable.

To be concrete, it is useful to consider a simple abstract example.
\Fig{abstract} shows the probabilistic graphical model for a set of
observations $\data_n$ taken some $N$ systems that are each parameterized by a
set of physical parameters $\local_n$ drawn from some global distribution
parameterized by the ``hyperparameters'' \hyper.
Note that each $\data_n$ and $\local_n$ are, in general, large vectors (or
blobs) with many entries.
Given this structure, we would like to compute the marginalized likelihood of
the full set of observations $\data = \{ \data_n \}$ given a specific
parameterization \hyper
\begin{eqnarray}
p(\data\,|\,\hyper) &=& \int \dd \local\,p(\data,\,\local\,|\,\hyper)
\nonumber\\
&=& \prod_{n=1}^N\int \dd \local_n\,p(\data_n,\,\local_n\,|\,\hyper) \quad.
\eqlabel{marg-like}
\end{eqnarray}
This integral isn't generally analytic but it could be solved using MCMC or
another approximate numerical technique.
A problem becomes apparent when we think a little harder about what might want
to \emph{do} with this marginalized likelihood.
In general, we will want to optimize it to find the maximum likelihood
population distribution of apply a prior function---parameterized by
\hyperhyper\ in \fig{abstract}---and draw posterior samples
$\hyper \sim p(\hyper\,|\,\data)$.
In both of these cases, we would have to perform the (probably extremely
expensive) integration in \eq{marg-like} some impractically huge number of
times.

It is possible, however, to greatly simplify this problem and make it
computationally tractable if we have a \emph{catalog}.
In this context, a catalog is a set of posterior probabilistic constraints on
the parameters $\local_n$ under some (known) \emph{interim priors}
$p(w_n\,|\,\hyper_0)$.
In particular, the best representation of these constraints is a set of
posterior samples $\local_n^{(j)} \sim p(\local_n\,|\,\data_n,\,\hyper_0)$.
Given a set of $J$ such samples, the integral in \eq{marg-like} can be
re-written as
\begin{eqnarray}
\int \dd \local_n\,p(\data_n,\,\local_n\,|\,\hyper) &=&
\int \dd \local_n\,p(\local_n\,|\,\hyper)\,p(\data_n\,|\,\local_n,\,\hyper)\,
    \frac{p(\local_n\,|\,\data_n,\,\hyper_0)}
         {p(\local_n\,|\,\data_n,\,\hyper_0)} \nonumber\\
&=& Z_n\,\int \dd\local_n\,
    \frac{p(\local_n\,|\,\hyper)}{p(\local_n\,|\,\hyper_0)} \,
    \frac{p(\data_n\,|\,\local_n,\,\hyper)}{p(\data_n\,|\,\local_n,\,\hyper_0)}
    \,p(\local_n\,|\,\data_n,\,\hyper_0) \nonumber\\
&\approx& \frac{Z_n}{J} \, \sum_{j=1}^J
    \frac{p(\local_n^{(j)}\,|\,\hyper)}{p(\local_n^{(j)}\,|\,\hyper_0)} \,
    \frac{p(\data_n\,|\,\local_n^{(j)},\,\hyper)}
         {p(\data_n\,|\,\local_n^{(j)},\,\hyper_0)}
\eqlabel{full-import}
\end{eqnarray}
where
\begin{eqnarray}
Z_n &=& \int \dd\local_n\,p(\local_n,\,\data_n\,|\,\hyper_0)
\end{eqnarray}
is the ``evidence'' for the data under the interim priors.
For our purposes, $Z_n$ is a constant that we (thankfully) never need to
compute.

There is a further simplification that we can make in the case where the
likelihood function does not depend on the hyperparameters.
In this case, \eq{full-import} becomes
\begin{eqnarray}
\int \dd \local_n\,p(\data_n,\,\local_n\,|\,\hyper) &\propto&
\int \dd\local_n\,
    \frac{p(\local_n^{(j)}\,|\,\hyper)}{p(\local_n^{(j)}\,|\,\hyper_0)} \,
    p(\local_n\,|\,\data_n,\,\hyper_0) \nonumber\\
&\approx& \frac{1}{J} \, \sum_{j=1}^J
    \frac{p(\local_n^{(j)}\,|\,\hyper)}{p(\local_n^{(j)}\,|\,\hyper_0)} \quad.
\end{eqnarray}
This simplification won't always be completely applicable and we'll come back
to it when we discuss the effects of survey selection functions below but most
of the expensive computations can normally be cancelled.

\begin{figure}[htbp]
\begin{center}
    \includegraphics{abstract.pdf}
\end{center}
\caption{%
\figlabel{abstract}}
\end{figure}

In this \paper, we will focus on the problem of inferring the population
parameters from a noisy incomplete \emph{catalog}.
This catalog is defined as some set of posterior probability
functions---preferably represented as samples---for some ``observable'' set of
systems.

\section{The model}

\paragraph{Selection function}
We assume that the selection function is specified by a probability
distribution function that gives the probability of observing a transit with a
given signal-to-noise ratio.
Although we don't perform the integral explicitly, this assumption requires a
marginalization over the period and inclination distribution.
Instead, we choose to model the selection function as a scaled logistic
function of signal-to-noise \sn:
\begin{eqnarray}
p(\isobs=1\,|\,\sn,\,\selection) &=&
\frac{\Pmax}{1+\exp \left [ -(\sn-\snmin) / \selectwidth^2 \right ]} \quad.
\end{eqnarray}
Particular period and inclination distributions imply different values for
\Pmax.
